% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Dummy Chapter:
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% The Introduction:
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\fancychapter{Minimal models of electron correlations in energy bands}
\label{cap:hubbard}

\slshape

The interactions between the electrons in a solid give rise to effects that arise specifically due to the many-body nature of the system.
The Hubbard model is a minimal model that encapsulates electron correlations.
It goes beyond the periodic ionic potential perturbation to the free electron gas or tight binding approaches, which lead to band theory.
One obtains the Hubbard Hamiltonian by adding the simplest possible electron-electron interaction term to a tight binding Hamiltonian: an on-site interaction term that penalizes double occupancy.
From it, we can make predictions about properties of a strongly correlated system, namely magnetic and superconducting behavior, and metal-insulator transitions.
We start with an overview of the Hubbard model and in the following chapters we provide details on how to simulate it numerically using Quantum Monte Carlo. We discuss the original motivation provided by Hubbard to introduce the model and show how its Hamiltonian arises as an approximate representation of the Coulomb repulsion between electrons in an energy band. Then, we present exact solutions for particular limiting cases, which will be used to crosscheck our simulations. In particular, we show that, in the limit where the interaction is large, the effective Hamiltonian at half filling corresponds to an atomic Heisenberg model defined in the appropriate Hilbert space with one electron per site.
Then, we introduce Green's functions and explain how they relate to Wick's theorem. We proceed by formulating the mean field theory of the Hubbard model, and comparing results for a \acs{1D} chain with those obtained from Stoner's criterion.
Finally, we show how one can generalize the Hubbard Hamiltonian and finish by discussing what type of interaction terms may be recast in a way that is more prone to simulation by writing the problem in terms of free fermions interacting with an external field.

\normalfont

\input{ch-hubbard/modelling.tex}
\input{ch-hubbard/hubbard.tex}
\input{ch-hubbard/mott-insulators.tex}
\input{ch-hubbard/exact-solutions.tex}
\input{ch-hubbard/effective-heisenberg.tex}
\input{ch-hubbard/greens-functions.tex}
\input{ch-hubbard/magnetism-mft.tex}
\input{ch-hubbard/variants.tex}
\cleardoublepage
