% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Dummy Chapter:
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% The Introduction:
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\fancychapter{Finite Temperature Auxiliary Field Quantum Monte Carlo}
\label{cap:finiteTafqmc}

We will now discuss a numerical method to simulate the Hubbard model. Among the various methods belonging to the family of QMC methods, auxiliary field QMC allows us to circumvent the sign problem for the half filled Hubbard model. The sign problem is an uncontrolled numerical error due to the antisymmetry of the many-electron wave function, leading to oscillations in the sign of the quantities that we are interested in measuring. These oscillations deem the algorithm exponentially complex in the size of the system, in general, but it possible to overcome this hurdle for a class of models, namely the Hubbard model at half filling. The difficulty lies in computing averages of quantities that are very close to zero, on average, but have a large variance, i.e. $\sigma_X / \left\langle X \right\rangle \gg 1$.

We seek a computable approximation of the projection operator $\mathcal{P}$ defined in equation (\ref{eq:projection}). As we shall see, it is found by using a discrete Hubbard-Stratonovich transformation. This transformation introduces an auxiliary field (consisting basically of Ising spins), and we use Monte Carlo to sample configurations from the distribution corresponding to this \emph{classical} configuration space.

For now, let us assume half filling $\mu = 0$, so that there is no sign problem. In fact, many interesting phenomena occur at half filling, for example magnetic ordering and the Mott metal-insulator transition.

\input{3.GSafqmc/1.sectiona.tex}
\input{3.GSafqmc/2.sectionb.tex}
\cleardoublepage
