
\fancychapter{Introduction}
\label{cap:int}

\slshape

The isolation of graphene in 2004 has led to a growing interest of the scientific community in two-dimensional materials revealing extraordinary properties.
In fact, their very existence was not expected a priori because at first sight they seem to violate the Mermin-Wagner-Hohenberg theorem \cite{mermin_absence_1966, coleman_there_1973, hohenberg_existence_1967}, a no-go theorem that forbids ordering below three dimensions at finite temperature.
Since graphene was discovered, a plethora of similarly stable \ac{2D} materials has been discovered.
A vast set of open problems remains to be solved within the realm of their fascinating and counterintuitive properties. These are often tackled by carrying out numerical simulations. \ac{QMC} is a simulation method that is amply applicable to condensed matter physics problems. Despite the system size being constrained due to limited simulation time, reliable and accurate solutions are provided to the otherwise intractable quantum many-body problem. The field is currently very active and method optimization can prove crucial in applications to real physical systems. In this work, \acs{QMC} is used to simulate a two-dimensional system with strong electron interactions giving rise to promising properties: a nanostructure made of a recent member of the \acs{2D} materials family called a nanoribbon.

\normalfont


\input{ch-introduction/1.strongly_correlated.tex}
\input{ch-introduction/2.state.tex}
\input{ch-introduction/3.contributions.tex}
\input{ch-introduction/4.outline.tex}

\cleardoublepage