% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Dummy Chapter:
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% The Introduction:
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\fancychapter{Applications and original results}
\label{cap:applications}

\slshape

We simulate the Hubbard model on the \acs{1D} chain and on the \acs{2D} square lattice, for which results are well established from either an analytical or numerical perspective.
Then, we consider the honeycomb lattice with  nanoribbon boundary conditions, and reproduce some results in the literature.
We compare the obtained results and the run time with \texttt{QUEST}, the flagship implementation of auxiliary field \acs{QMC}.
Finally, we carry out original calculations, applying our code to a minimal model of \acs{TMD} nanoribbons.
We compare our \acs{QMC} results with those obtained by us in the mean field (MF)  approximation.
The latter is formulated allowing a site and orbital-dependent mean field, and the self-consistency relation is solved iteratively.
We characterize the convergence issues arising in this iterative procedure.
Our original MF results suggest that in the model we consider \acs{TMD} nanoribbons could host edge-magnetism at finite temperature.
The obtained phase diagram shows transitions between non-magnetized and two distinct edge-magnetized configurations.
We use our unbiased \acs{QMC} simulations to further probe the system, searching for these or other eventual phase transitions not captured by mean field theory.

\normalfont

\input{ch-applications/1d-chain.tex}
\input{ch-applications/square-lattice.tex}
\input{ch-applications/nanoribbon.tex}
%\cleardoublepage
